---
title: "PA1_template.Rmd"
author: "John Shomaker"
date: "June 4, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Reproducible Research: Assignment 1

## Activity Monitoring Data 

The variables included in this dataset are:

**steps:** Number of steps taking in a 5-minute interval (missing values are coded as NA)  
**date:** The date on which the measurement was taken in YYYY-MM-DD format   
**interval:** Identifier for the 5-minute interval in which measurement was taken  

The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.

## 1. Data Import & Cleaning

```{r download}
## Install packages and libraries
## Download the activity.csv file and read to dataframe; initially omit NAs
## Convert 'date' column to date class; interval to factor

## install.packages("data.table")
## library(data.table)

filename <- "activity.csv"

if (!file.exists(filename)){
      fileURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
      download.file(fileURL, filename, method="curl")
}  

activity_all <- read.csv(filename, stringsAsFactors = FALSE, na.strings = "NA")
activity <- na.omit(activity_all)

activity$date <- as.Date(activity$date, "%m/%d/%y")
activity$interval <- factor(activity$interval)

options(scipen=999)

```

## 2. Total Steps Taken Each Day (Histogram)

```{r totalsteps}
## Summarize the total steps taken each day
## Create a histogram (frequency) of daily step count

activity_sum <- aggregate(steps ~ date, data = activity, sum)
hist(activity_sum$steps,
     main = "Activity-Total Daily Steps",
     xlab = "Steps Taken per Day", ylab = "Frequency (# of Days)")

```

## 3. Mean/Median Steps Taken Each Day

NOTE: Technically, it's the mean/median of the daily step totals

``` {r mean_median}

## Compute mean and median for the activity_sum (by day created in question #2)

activity_mean <- round(mean(activity_sum$steps), digits = 2)
activity_median <- median(activity_sum$steps)

```
The mean number of steps taken per day (across the sample number of days) is `r activity_mean`.
And, the median number of steps taken per day is `r activity_median`.


## 4. Time Series of Average Steps Taken

NOTE: Question is unclear, but given data set interpreted as a plot of the 
average (mean) steps per interval (across days), plotted by interval

``` {r avg_steps_interval}

## Find the average steps per interval across days
## Plot a line chart with intervals on x-axis

interval_avg <- aggregate(steps ~ interval, data = activity, mean)

plot(interval_avg$interval, interval_avg$steps, type = "l",
     xlab = "Daily 5-Minute Intervals", ylab = "Avg. Steps per 5-Min Interval",
     main = "Average Steps By Day Interval")

```

## 5. 5-Minute Interval with Greatest Average Steps

``` {r max_interval}

## Find the interval with the largest average number of steps
## and present the interval number (factor) and steps amount

interval_sort <- interval_avg[order(interval_avg$steps), ]
interval_max <- interval_sort$interval[nrow(interval_sort)]
interval_max_amt <- round(interval_sort$steps[nrow(interval_sort)], digits = 2)

```
The 5-minutes interval with the greatest average steps is `r interval_max`, 
which, on average, has `r interval_max_amt` steps.  


## 6. Imputing Data for Missing Data

Replace NA rows with average steps for the particular interval

```{r impute}

## Put missing rows in a separate dataframe
## Impute the missing steps value based on interval from interval_avg
## Insert the lookup value (via merge) and clean up table
## Show abbreviated head & tail of imputed table

missing <- subset(activity_all, is.na(steps))
imputed <- merge(missing, interval_avg, by="interval")
imputed <- imputed[, -2]
colnames(imputed)[3] <- "steps"

print(head(imputed, 6))
print(tail(imputed, 6))

```

## 7. Total Steps Taken Each Day (Histogram) with Imputed Data

```{r totalsteps_plus_imputed}

## Bind the activity df (that excluded NAs) + the NA records (with imputed)
## First make sure class types and column positions match
## Recreate the histogram

imputed$date <- as.Date(imputed$date, "%m/%d/%y")
imputed$interval <- factor(imputed$interval)
imputed$steps <- as.integer(imputed$steps)

activity_temp <- activity[, c(2,3,1)]
imputed_temp <- imputed[, c(2,1,3)]

act_imp <- rbind(activity_temp, imputed_temp)

act_imp_sum <- aggregate(steps ~ date, data = act_imp, sum)
hist(act_imp_sum$steps,
     main = "Activity-Total Daily Steps (with Imputed Records)",
     xlab = "Steps Taken per Day", ylab = "Frequency (# of Days)")

```

## 8. Panel Plot Comparing Interval Steps - Weekdays vs. Weekends

```{r weekdays_weekends}

## Subset the activity data (with imputed) for weekdays and weekends
## Searching for the first letter being an "S" is a short-cut
## For each subset find the mean number of steps by interval
## Merge the interval analyses to simplify plotting

act_wday <- subset(act_imp, substring(weekdays(act_imp$date), 1, 1) != "S") 
act_wend <- subset(act_imp, substring(weekdays(act_imp$date), 1, 1) == "S")

wday_interval_avg <- aggregate(steps ~ interval, data = act_wday, mean)
wend_interval_avg <- aggregate(steps ~ interval, data = act_wend, mean)

wday_wend_avg <- merge(wday_interval_avg, wend_interval_avg, by = "interval")
colnames(wday_wend_avg)[2] <- "weekday_steps"
colnames(wday_wend_avg)[3] <- "weekend_steps"

par(mfrow = c(1, 2))

plot(wday_wend_avg$interval, wday_wend_avg$weekday_steps, type = "l",
     xlab = "Daily 5-Minute Intervals", ylab = "Avg. Steps per 5-Min Interval",
     main = "Weekday Steps By Interval")

plot(wday_wend_avg$interval, wday_wend_avg$weekend_steps, type = "l",
     xlab = "Daily 5-Minute Intervals", ylab = "Avg. Steps per 5-Min Interval",
     main = "Weekend Steps By Interval")

```

